Verification of Stencil Computations with Spatial Specifications

### 2016-11-12 paper.tex
reworking the checking procedure

\section{Checking and inference}

%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
\newcommand{\cons}[3]{#2 \;\, \textbf{cons} \,\, #1}
\newcommand{\consA}[3]{#2 \;\, \textbf{cons}_M \,\, #1}

\subsection{Specification checking}

\begin{definition}
Consistency is defined in terms of the binary predicate
$\textbf{cons}$ in Figure~\ref{fig:consistency}.
Consistency relates models of source code and models 
of specifications (which are bounded), thus
 $\textbf{cons} \subseteq \mathcal{P}(\domainVal{}^n)
\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n))$. 
We denote sets of indexing schemes $\mathcal{P}(\domainVal{}^n)$
that come from a specification model by $S$ and from source code model
by $C$.

@@

Rules $(1)$ and $(*)$ extend the $\textbf{cons}$
predicate to $\textbf{cons}_M \subseteq
\textsf{Mult}(\mathcal{P}(\domainVal{}^n)) 
\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
which inludes multiplicity information. Both rules 
state that consistent models remain consistent if they
have the same multiplicity. Relatedly, rule $(1 < *)$ ensures
that absence of a \texttt{readOnce} modifier in the specification
does not prohibit a set of unique stencil accesses in the
code from being consistent.
\end{definition}

\begin{figure}[t]
\begin{align*}
\begin{array}{c}
%%
\trule{up}\dfrac{\forall u \in C \, \exists v \in S .  \, u \preceq v}
      {\cons{\textsf{upper}(S)}{C}{n}}
\qquad
\trule{low}\dfrac{\forall v \in S \, \exists u \in C . \, u \preceq v}
      {\cons{\textsf{lower}(S)}{C}{n}} \\[1.5em]
%%
\trule{exact}\dfrac{\cons{\textsf{lower}(S)}{C}{n} \qquad \cons{\textsf{upper}(S)}{C}{n}}
  {\cons{\textsf{exact}(S)}{C}{n}} \\[1.5em]
%%
\trule{both}\dfrac{\cons{\textsf{lower}(S_1)}{C}{n} \qquad \cons{\textsf{upper}(S_2)}{C}{n}}
  {\cons{\textsf{both}(S_1, S_2)}{C}{n}}\\[1.5em]
%%
\trule{$1$}
\dfrac{\cons{S}{C}{n}}{\consA{\textsf{once}(S)}{\textsf{once}(C)}{n}}
\;\;
\trule{$*$}
\dfrac{\cons{S}{C}{n}}{\consA{\textsf{mult}(S)}{\textsf{mult}(C)}{n}}
  \\[1.5em]
\trule{$1 < *$}
\dfrac{\cons{S}{C}{n}}{\consA{\textsf{mult}(S)}{\textsf{once}(C)}{n}}
\end{array}
\end{align*}
\caption{Consistency between models of code and specification}
\label{fig:consistency}
\vspace{-0.9em}
\end{figure}

>>>>

\section{Analysis, checking, and inference}
\label{sec:analysis}

%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
\newcommand{\consName}{\textbf{cons0}}
\newcommand{\cons}[2]{#2 \;\, \consName{} \,\, #1}
\newcommand{\consAName}{\textbf{cons}}
\newcommand{\consA}[2]{#2 \;\, \consAName{} \,\, #1}

\noindent
We outline here the procecures for checking the conformance
of source code against specifications (\S~\ref{subsec:checking}) 
and for infering specifications from code (\S~\ref{subsec:inference}).
Both rely on the analysis of source code, converting array indexing
expressions into sets of index schemes. We outline this analysis
first (\S~\ref{subsec:analysis}). Note that the analysis procedure
can be made arbitrarily more complex and wide-ranging independent
of the checking and inference procedures.

\subsection{Analysis of array-based code}
\label{subsec:analysis}

\newcommand{\neigh}{\textsf{neigh}}

\paragraph{Syntax helper functions}
The function \rhsExp{} maps source statements to a
set of expressions which occur in right-hand positions (\ie{}, not the
target of an assignment). The function \var{} maps expressions to a
set of the variables used in right-hand positions.

\paragraph{Analysis pre-requisites}
The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG);
\textbf{(2)} induction variables per basic block;
\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathsf{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\textbf{(4)} type information per variable, where we use the predicate
\arrayTy{} to classify variables of array type.

\paragraph{Core analysis}
For each assignment statement \textit{assg} within a loop whose left-hand side is an array
subscript on neighbourhood indices, a finite map is computed mapping
array variables to \emph{multisets} of indexing schemes.
This is captured by the function $\textit{analyse}$:
%
\begin{align*}
\hspace{-0.5em}
\begin{array}{lr}
\textit{analyse}(a(\overline{e_1}) = e_2)
 := \qquad\qquad\qquad & \textit{where} \; \neigh(\overline{e_1}) \\[0.3em]
\multicolumn{2}{l}{\qquad \bigcup\{a' \mapsto \{
\textsf{schematic}(\bar{e})\} \mid a'(\bar{e}) \leftarrow \mathsf{flowsTo}(e_2),
  \arrayTy(a')\}}
\end{array}
\end{align*}
%
The predicate $\neigh$ classifies neighbourhood indices, based on
the induction variables of the loop (implicit here).
Thus, we focus on assignments to an array subscript where the LHS 
index $\overline{e_1}$ is a neighbourhood index.  For
all array subscripts that flow to the right-hand side of this
statement, a finite map is constructed, mapping each array variable
to a \emph{multiset} of indexing schemes for its subscripts, computed
with \textsf{schematic} (Defn.~\ref{def:schematic},
p.~\pageref{def:schematic}).

Note that \textsf{schematic} is undefined ($\bot$) if
$\bar{e}$ contains relative indices which are not neighbourhood
indices. The \textit{analysis} function feeds into the
\textit{modelise} function which filters out those arrays 
which contain undefined indexing schemes and adds multiplicity
information:
%
\begin{align*}
& \mathit{modelise} (\textit{assg}) 
= a \mapsto 
\begin{cases}
\bot & \bot \in M \\
\textsf{only}(|M|) & \forall u \in M . \#(M, u) = 1 \\
\textsf{mult}(|M|) & \forall u \in M . \#(M, u) = n
\end{cases} \\
& \quad \textit{where} \; M = \mathit{analyse}(\textit{assg}) (a)
\end{align*}
where $\#(M, u)$ denotes the multiplicity of an element $u$ of the
multiset $M$ and $|M|$ denotes the set of elements in the multiset
$M$.

Thus, the analysis maps source code to index schemes, which is then
augmented with multiplicity information by $\mathit{modelise}$,
\ie{} values in $\mathsf{Mult}(\mathcal{P}((\domainVal{})^n))$.

If the LHS $\overline{e_1}$ contains non-$0$ neighbour offsets 
then all RHS schemes are \emph{relativised} by
$\textsf{schematic}(\overline{e_1})$. For example, 
if \texttt{a(i+1) = b(i)} then relativisation of the RHS by the LHS
produces an analysis which is equivalent to the analysis of \texttt{a(i) = b(i-1)}. 
\dnote{Somewhere we need a discussion about syntactic vs semantic
analysis...}

\begin{example}
We demonstrate the analysis and modelise procedures over the
following example program which computes the mean value
of a five-point stencil:
%%
\begin{minted}{fortran}
do i = 1, (n-1)
  do j = 1, (m-1)
    x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
    b(i, j) = (a(i, j) + x + y) / 5.0
  end do
end do
\end{minted}
%
The \textit{analyse} function matches on
line 4, with the following set for $\textsf{flowsTo}(\texttt{a(i, j) + x +
  y})$:
%
\begin{align*}
\{\texttt{a(i-1, j)}, \texttt{a(i+1, j)}, \texttt{a(i, j-1)},
  \texttt{a(i, j+1)}, \texttt{a(i, j)}\}
\end{align*}
Subsequently the result of \textit{analyse} on line 4 yields the map:
\begin{align*}
\texttt{a} \mapsto \{\vtwoh{-1}{0}^1, \vtwoh{1}{0}^1,
          \vtwoh{0}{-1}^1, \vtwoh{0}{1}^1, \vtwoh{0}{0}^1\}
\end{align*}
%
where multiplicities are denoted by superscripts. The \emph{modelise}
function then yields the following since all multiplicities in the model are one:
%%
\begin{align*}
\texttt{a} \mapsto \mathsf{only}(\{\vtwoh{-1}{0}, \vtwoh{1}{0},
          \vtwoh{0}{-1}, \vtwoh{0}{1}, \vtwoh{0}{0}\})
\end{align*}
%%
\end{example}

%Given an assignment $\textit{assg}$ to an array subscript in the source language
%and an array variable $a$, then $\mathit{modelise}(\textit{assg}, a)$ performs
%an analysis to find all array subscripts of $a$ converting them to
%index schemes. For the moment, we leave this analysis abstract
%as the function $\emph{analyse}$ in the following definition of $\emph{modelise}$:
%

\subsection{Specification checking}
\label{subsec:checking}

In checking mode, the access pattern of stencil computations in the
source language is verified against any specification associated with
them in the code. The model of
\S~\ref{sec:semantics} guides this process.  Checking
proceeds by generating a model from a specification and generating
a model from the source code, comparing them for
\emph{consistency}. The notion of consistency is built out of index scheme 
\emph{generalisation} (Sect.~\ref{subsec:schemes-and-gen}).

\begin{definition}[Checking]
Let $a$ be an array variable and let
$S \in \textit{spec}$ be a specification associated to $a$ 
via \texttt{!= stencil $S$ :: a}. A stencil kernel 
 $v(\overline{e_1}) = e_2$ in the source language
conforms to $S$ if the predicate $\textbf{check}_n(S, v(\overline{e_1}) = e_2, a)$ holds, 
defined:
\begin{align*}
& \textbf{check}_n(S, \textit{assg}, a) =
  \consA_n{\interp{S}_n}{\textit{modelise}(\textit{assg})(a)}
&
\end{align*}
Thus, checking proceeds by using \textit{modelise} 
(\S~\ref{subsec:analyis}) to extract index schemes from source code
augmented with multiplicitly for the whole set of schemes. 
This is compared with the model of the specification $\interp{S}_n$ by
the consistency predicate $\consAName$.
\end{definition}

\begin{definition}
Consistency relates models of source code to models 
of specifications (which are bounded) via two predicates:
\begin{align*}
\consName_n & \subseteq \mathcal{P}(\domainVal{}^n)
\times \textsf{Approx}(\mathcal{P}(\domainVal{}^n)) \\
\consAName_n & \subseteq \textsf{Mult}(\mathcal{P}(\domainVal{}^n))
\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))
\end{align*}
%
where $\consName$ relates models without multiplicity.
Figure~\ref{fig:consistency} defines the two predicates, where for
brevity we make implicit the subscript $n$ defining the
dimensionality.  We denote sets of indexing schemes
$\mathcal{P}((\domainVal{})^n)$ that come from a specification model
by $M$ and from source code model by $N$.

@@

Rules $(1)$ and $(*)$ extend the $\textbf{cons}$
predicate to $\textbf{cons}_m \subseteq
\textsf{Mult}(\mathcal{P}(\domainVal{}^n)) 
\times \textsf{Mult}(\textsf{Approx}(\mathcal{P}(\domainVal{}^n)))$
which inludes multiplicity information. Both rules 
state that consistent models remain consistent if they
have the same multiplicity. Relatedly, rule $(1 < *)$ ensures
that absence of a \texttt{readOnce} modifier in the specification
does not prohibit a set of unique stencil accesses in the
code from being consistent.
\end{definition}

\begin{figure}[t]
\begin{align*}
\begin{array}{c}
%%
\trule{up}\dfrac{\forall u \in M \, \exists v \in N .  \, u \preceq v}
      {\cons{\textsf{upper}(N)}{M}}
\qquad
\trule{low}\dfrac{\forall v \in N \, \exists u \in M . \, u \preceq v}
      {\cons{\textsf{lower}(N)}{M}} \\[1.5em]
%%
\trule{exact}\dfrac{\cons{\textsf{lower}(N)}{M} \qquad \cons{\textsf{upper}(N)}{M}}
  {\cons{\textsf{exact}(N)}{M}} \\[1.5em]
%%
\trule{both}\dfrac{\cons{\textsf{lower}(M_1)}{M} \qquad \cons{\textsf{upper}(M_2)}{M}}
  {\cons{\textsf{both}(M_1, M_2)}{M}}\\[1.5em]
%%
\trule{$1$}
\dfrac{\cons{N}{M}}{\consA{\textsf{once}(N)}{\textsf{once}(M)}}
\;\;
\trule{$*$}
\dfrac{\cons{N}{M}}{\consA{\textsf{mult}(N)}{\textsf{mult}(M)}}
  \\[1.5em]
\trule{$1 < *$}
\dfrac{\cons{N}{M}}{\consA{\textsf{mult}(N)}{\textsf{once}(M)}}
\end{array}
\end{align*}
\caption{Consistency between models of code and specification}
\label{fig:consistency}
\vspace{-0.9em}
\end{figure}

\begin{example}
  We give an extended example of verifying a stencil computation
  against its specification. We consider a more general, but still
  consistent, version of the \textit{five-point} stencil specification
  inferred previously in \S~\ref{subsec:inference}:
%
\begin{minted}[breakindent=8.1em]{fortran}
!= region :: fivepoint = centered(depth=1,dim=1) + centered(depth=1,dim=2)
!= stencil readOnce, fivepoint :: b
a(i,j) = b(i,j-1)+b(i,j)+b(i,j+1)+b(i-1,j)+b(i+1,j)
\end{minted}
%
We refer to the assignment on line three as $l3$ for brevity.

Verification of the code against the spec is then given by 
$check(({\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}),l3,\texttt{b})$. 
We explain the steps.
\begin{enumerate}  
%
\item Generate the model from the specification as in
  \S~\ref{sec:semantics}. 
%
\begin{align*}
 N_s= \{  \vtwo{-1}{\infty},
          \vtwo{0}{\infty},
          \vtwo{1}{\infty},
          \vtwo{\infty}{-1},
          \vtwo{\infty}{0},
        \vtwo{\infty}{1} \}  \\[0.4em]
\interp{{\textcolor{darkgreen}{\texttt{readOnce, fivepoint}}}}_2 =
  \textsf{once}(\textsf{exact} (N_s))
\end{align*}
%
\item Obtained the code model via $\textit{modelise}$, giving:
%
\begin{align*}
M_c = \{
    \vtwo{0}{-1},
    \vtwo{0}{0},
    \vtwo{0}{1},
    \vtwo{-1}{0},
    \vtwo{1}{0}
  \}
\\[0.4em]
\textit{modelise}(l3, \texttt{b}) = \textsf{once} (M_c)
\end{align*}
%
\item Consistency $cons_2(\textsf{once}(\textsf{exact}(N_s)),
  \textsf{once}(M_c))$ holds when 
  $\textit{cons}_2(\textsf{mult}(\textsf{exact}(N_s)),\textsf{mult}(M_c))$ (Fig.~\ref{fig:consistency}) 
  which holds when: 
  \begin{align*}
     & \textit{cons}_2(\textsf{mult}(\textsf{up}(N_s)),\textsf{mult}(M_c)) \\
   \wedge \; & \textit{cons}_2(\textsf{mult}(\textsf{low}(N_s),\textsf{mult}(M_c))
  \end{align*}
%
\item This condition holds by the following sets of scheme
  generalisations: for the upper bound, $\forall u \in M_c,
  \exists v \in N_s . u \preceq_2 v$: 
  \begin{align}
    \begin{array}{l}
     \vtwoh{0}{-1} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{0}{1} \preceq_2 \vtwoh{0}{\infty} \\[0.5em]
     \vtwoh{-1}{0} \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{\infty}{0}
     \end{array}
  \label{eq:check-example-upper}
  \end{align}
  % 
  and for the lower bound: $\forall v \in N_s, \exists u \in M_c . u
  \preceq_n v$: 
  \begin{align}
    \hspace{-1em}
    \begin{array}{l}
     \vtwoh{-1}{0} \preceq_2 \vtwoh{-1}{\infty} \quad
     \vtwoh{0}{0} \preceq_2 \vtwoh{0}{\infty} \quad
     \vtwoh{1}{0} \preceq_2 \vtwoh{1}{\infty}  \\
     \vtwoh{0}{-1}  \preceq_2 \vtwoh{\infty}{-1} \quad
     \vtwoh{0}{0}  \preceq_2 \vtwoh{\infty}{0} \quad
     \vtwoh{0}{1}  \preceq_2 \vtwoh{\infty}{1}
     \end{array}
    \label{eq:check-example-lower}
  \end{align}
  Since both the consistency predicate is satisfied, the array
  accesses flowing to line three in the stencil computation are valid
  with respect to the {\textcolor{darkgreen}{\texttt{fivepoint}}} specification.
\end{enumerate}
\vspace{-0.7em}
\end{example}
\begin{example} 
We consider now an erroneous user modification to $l3$ with an
additional term \texttt{b(i+2,j+2)} such that $l3'$ is:
%
\begin{minted}[firstnumber=3]{fortran}
a(i,j) = b(i,j-1) + b(i,j) + b(i,j+1) + b(i-1,j) + b(i+1,j) + b(i+2,j+2)
\end{minted}
%
The checking procedure proceeds in the same way as the first three
steps of the previous example, except that the result of
$\mathit{modelise}(l3',\texttt{b})$ generates an additional scheme
$\vtwoh{2}{2}$. Since the spec model has not changed and the code
model is has only the additional scheme $\vtwoh{2}{2}$, the generalisations for
the lower bound \eqref{eq:check-example-lower} 
of step 4 above still holds since its consistency condition
universally quantifies over the \emph{spec model}. However, for
the upper bound, the consistency condition universally quantifies
over scheme $u$ of the \emph{code model} for which there exists
a scheme $v \in N_s$ such that $u \preceq_2 v$. Thus, we require a $v
\in N_s$ such that $\vtwoh{2}{2} \preceq_2 v$ but no such generalising
scheme existing in the spec model. Hence the model fails to hold as an upper bound.

Note that if the specification were as follows:
%
\begin{minted}{fortran}
!= stencil atLeast, readOnce, fivepoint :: b
\end{minted}
%
then consistency holds as it no longer requires the failing upper-bound case 
$\mathit{cons}(\textsf{mult}(\textsf{up}(N_s)),\textsf{mult}(M_c))$.
 This makes sense semantically since the region ${\textcolor{darkgreen}{\texttt{fivepoint}}}$
is indeed contained in the regions touched by $\texttt{b}$ in $l3'$. 
\end{example}

\dnote{Build this into the definitions earlier I think}
One way the implementation may deviate from the model is elimination of all $\infty$
offset vector that occurs when all indices are in non-neighbourhood form \eg{}
\mintinline{fortran}{b(42,0)}. The only generalisation scheme that can unify
with this pattern is the most general one, meaning it requires outright
rejection of an otherwise fine stencil computation. See the following example,
where not discarding all $\infty$ access pattern prevents verification:
%
\begin{minted}{fortran}
!= stencil forward(dim=1,depth=1) :: a
a(i) = a(i) + a(i+1) + a(1)
\end{minted}
